Predicting the conformational preferences of proteins using a physics-based free energy method

### 2013-05-05 Paper/paper.tex
Revisions on manuscript and SI may 5, Alberto

Calculation of free energy differences is of central importance in the simulation of biochemical
systems. It is particularly difficult to calculate between pairs of macromolecular conformations as
well as a computationally expensive task with existing methods. In this work, confinement approach
is used to calculate absolute free energies of biomolecular systems. This method provides two main
advantages: it does not require a reaction coordinate or transition path and it is fast to compute.
Free energy calculated can be decomposed into a per residue contribution in an approximate way. Per
residue free energy allows us to identify the reason behind conformational preferences in
biomolecules. Through out the article we show its use in different challenging modeling problems. In
particular, we show its use in predicting the conformational preference of chamaleon sequences
(sequences with high sequence identity and different folds). This sequence dependent conformational
preferences and per residue free energy decomposition set the stage for the use of this method in
protein design.

@@

\begin{table}
\begin{center}
\caption{The confinement method assigns a more favorable free energy to the experimentally determined structure than to
    computer-generated predictions. For each target, we examined as many as five predictions submitted by CASP
    participants. We report the free energy difference between the most favorable decoy and the experimentally
    determined structure. Positive $\Delta\Delta G (=G_{best~decoy} - G_{native})$ values indicate that the experimental structure is predicted to be
    more favorable than any of the decoys.}
\label{table:casp_control}
\begin{tabular}{l l l l}\hline
    CASP Target  & PDB Identifier & $\Delta \Delta G$ (kcal/mol) & GDT-TS \\ \hline
     T0531       &    2KJX        &          $11.15 \pm 0.70$    &  $44$ \\ \hline
     T0538       &    2L09        &          $-3.00 \pm 0.47$    &  $96$ \\ \hline
     T0540       &    3MX7        &          $16.94 \pm 0.49$    &  $70$ \\ \hline
     T0559       &    2L01        &          $2.10 \pm 0.24$     &  $94$ \\ \hline
     T0560       &    2L02        &          $18.00 \pm 0.49$    &  $94$ \\ \hline
     T0569       &    2KYW        &          $20.01 \pm 0.69$    &  $78$ \\ \hline
\end{tabular}
\end{center}
\end{table}

@@

We also tested the confinement method predictions for difference free energies on the chameleon sequences of Orban et
al.\cite{Alexander2007,He2008,Alexander2009,Bryan2010,He2012}. These are instances in which two highly
similar sequences fold into remarkably different structures. Orban and co-workers have designed a protein-G-like
sequence of 56-residues that is marginally stable in one of two possible folds. By mutating key residues in this
sequence they are able to stabilize one fold or the other (see Figure~\ref{fig:orban}). We refer to the $4\beta +
\alpha$ structure as the $\beta$ conformation, and we refer to the $3\alpha$ structure as the $\alpha$ conformation. One
pair of sequences (GA88/GB88) is 88 percent identical in sequence, differing in seven positions. Another pair
(GA95/GB95) is 95 percent identical, differing in three positions. Accurately predicting the structural preferences of
these structures presents a serious challenge for computational
methods \cite{Allison2011}.

@@

As is common in the CASP experiment, we assess our results in terms of Global Distance Test Total Score (GDT-TS)
\cite{Zemla2003}, which is a C$\alpha$ based measure of structural accuracy. It can be understood roughly as the
percentage of residues that are correctly positioned in the model (range 0 to 100, higher is better). We did have enough
computer model to analyze every model, so the initial models for the test were chosen from a selection of different
server groups that have done well in past CASP events.

@@

First, we examined the ability of the confinement method to rank different models for a target that have been generated
by a single group using the same methodology for all predictions. We examined two targets: T0559 and T0560 (see Table~\ref{table:casp_testcase}).

\begin{table}
\begin{center}
\caption{The six different CASP targets that are used in this study. The corresponding PDB Identifier and the 
         organism type are also listed.}
\label{table:casp_testcase}
\begin{tabular}{l l l}\hline
    CASP Target  & PDB Identifier &  organism type  \\ \hline
     T0531       &    2KJX        &  extracellular domain of the jumping  \\
                 &                &  translocation breakpoint protein  \\ \hline
     T0538       &    2L09        &  protein asr4154 from Nostoc sp. PCC7120     \\ \hline
     T0540       &    3MX7        &  human Fas apoptotic inhibitory molecule     \\ \hline
     T0559       &    2L01        &  protein BVU3908 from Bacteroides vulgatus   \\ \hline
     T0560       &    2L02        &  protein BT2368 from Bacteroides  \\ 
                 &                &  thetaiotaomicron         \\ \hline
     T0569       &    2KYW        &  domain of adhesion exoprotein from \\ 
                 &                &   Pediococcus pentosaceus \\ \hline
\end{tabular}
\end{center}
\end{table}

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=3.5 in]{T0540.pdf}
    \end{center}
    \caption{The confinement method can correctly rank the native structure and two models submitted by different
        prediction groups.}
\label{fig:T0540}
\end{figure}

\subsection*{The per-residue free energy is sensitive to small changes in protein conformation:}

@@

The confinement method clearly identifies two hydrophobic residues V59 and I61, which destabilize the predicted model
with respect to the experimental structure (Figure~\ref{fig:T0569_per_residue} and Figure S3). The sidechains of
these hydrophobic residues are oriented towards the protein hydrophobic core in the native NMR structure but oriented
towards the exterior of the protein and are solvent exposed in the model. These residues are part of a beta-sheet
in the experimental structure, but because of their sidechain orientation the corresponding beta-sheet become disordered in the predicted model 
(Figure~\ref{fig:T0569_per_residue} and Figure S4). There is also a large
difference around K76, which forms a salt-bridge with D11 in the predicted model, but not in the experimental structure.
This suggests that salt-bridge interactions are too favorable for the combination of force field and implicit solvent
model we use, which has been a problem noted in the past \cite{Roe2007}.

@@

Despite success for most of the studied systems, there are few failures, specially when the GDT-TS scores of the
compared structures are very close. One example is Target T0538, where we compared the experimental structure with three
models (Model 1: ``PconsR''---GDT-TS=96; Model 2: ``Shell''---GDT-TS=90; Model 3: ``FOLDIT''---GDT-TS=86). Contrary to
our expectation, the confinement method predicts that computer generated Model 1 is more stable than the crystal structure (Figure S5).
Per-residue free energy calculations (not shown) show that despite only small variations at the backbone level, the
sidechains are oriented in very different ways (Figure~\ref{fig:T0538compare}), giving rise to large differences in the stabilization of
certain residues. In particular, some of the differences arise from different salt bridge patterns and certain flexible
polar residues exposed to the surface. This unexpected result shows that the confinement method is very sensitive to
local interactions (including sidechain reorientation) and may indicate issues with the forcefield and implicit solvent
models used in the calculation.

\begin{figure}
\begin{center}
\includegraphics[width=4.9 in,height=4.0 in]{Target_538_compare.pdf}
\end{center}
\caption{The per residue free energy calculation found some differences of the sidechain orientation in computer generated model 1 and 
experimental structure of CASP target T0538. 
(A) Two salt bridges between residues E35-R32 and E28-K24 exist in the native structure. These are
absent in the computer generated model structure. Instead
(B) a new salt bridge is formed between residues R32-E28 in model. (C) Another salt bridge between
residues  R26-E50 in the experimental structure. (D) In comparison to the native 
structure, orientation of F51 differs in the model.}
\label{fig:T0538compare}
\end{figure}

@@

A part of the CASP experiment is dedicated to the quality assessment (QA) of predicted models \cite{Kryshtafovych2011}.
Here, predictors were asked to score each model on a scale (known as qmode) from 0 to 1, with higher values
corresponding to better models \cite{Kryshtafovych2011}. Most of the groups use consensus strategies in such experiments.
Here, we chose two computer generated models from CASP taget T0538, where the top performing group "MUFOLD-WQA" \cite{Wang2011} failed.
One of the model was from "PconsR" (GDT-TS = 83, qmode 1 = 0.5865) and another from "MULTICOM-NOVEL" ( GDT-TS = 83, qmode 1 = 0.5865). We 
choose these two models as the model by PconsR was the most accurate
predicted model for this target, whereas the other model was predicted best by QA test of MUFOLD-WQA. The calculated
free energy using confinement method indicate that the model by PconsR is more stable by $3.9 Kcal/Mol$ which support
the GDT\_TS trend.  Although consensus prediction strategies do well specially when dealing with large numbers of structures, they 
can miss a good answer that is only predicted by a few methods. At least in this case, the confinement
method is able to capture this structure that was otherwise missed.

@@

The confinement method has been described in details in ref. by Tyka et al. \cite{Tyka2006} and
Cecchini et  al. \cite{Cecchini2009}. The basic approach of the confinement approach is the same in
both these papers. A thermodynamic cycle is used to compute the free energy between conformations A
and B. However, there are some small technical differences between the two approaches. Here we briefly
describe the procedure that we used.

@@

\item  In the first step, a minimization of A and B conformations are performed. These minimized conformations
    ($A^{\ast}$ and $B^{\ast}$) are the reference conformation of that state.

\item The free energy of confining the ensemble (A or B) to a tighter ensemble ($A^\ast$ or $B^\ast$) is calculated.
    This is done by gradually confining each atom to its position in the reference conformation ($A^{\ast}$ or
    $B^{\ast}$) using a series of progressively stronger position restraints. This is done by running 21 molecular
    dynamics simulation (each 20 ns long) for each leg of the thermodynamic cycle, where the harmonic restraint force
    constant was scaled from 0.00005 Kcal/Mol (mostly free) to 81.92 kcal/mol (tightly restrained). The free energy for
    this step is estimated using a thermodynamic integration approach developed by Tyka et. al. \cite{Tyka2006}. The
    confinement free energies calculated this step are denoted as $\Delta G_{A,A^\ast}$ and $\Delta G_{B,B^\ast}$ in
    Figure~\ref{fig:method}.

@@

One advantage of confinement method is that
none of the simulation during restraining step depends on each other. Therefore, with available computer resources it can be fast 
to compute. The computational cost for a $56$-residue proteins with chameleon sequence that is reported in this article is only $4$ hours 
in a single gpu computer. All calculations were performed with the Amber 11 suit of programs~\cite{Case2012,Goetz2012} in combination with ff99SB
forcefield~\cite{Hornak2006} and generalized born implicit solvent \cite{Mongan2006}.

Interestingly, we extend the
method for calculation of per residue free energy in an approximate way. For this purpose, the confinement energy,
$\Delta G_{A,A*}$ and $\Delta G_{B,B*}$ of each residue is calculated in the usual numerical way as described in ref. by
Tyka et al. \cite{Tyka2006}. The internal energy of each residue is calculated using the decomp module of amber from the
final restrained trajectory. We call this method approximate as we ignore the entropic contribution from the normal mode
or quasiharmonic analysis. However, this contrbution to the total free energy is much smaller, which allow us to study
the mechanistic details of conformational preference of each residue.

>>>>

The calculation of free energy differences is of central importance in the simulation of biochemical
systems. It is particularly difficult to calculate between pairs of macromolecular conformations as
well as a computationally expensive task with existing methods. In this work, the confinement approach
is used to calculate absolute free energies of biomolecular systems. This method provides two main
advantages: it does not require a reaction coordinate or transition path and it is fast to compute.
The calculated free energy can be decomposed into a per residue contribution in an approximate way. Per
residue free energy allows us to identify the reason behind conformational preferences in
biomolecules. Through out the article we show its use in different challenging modeling problems. In
particular, we show its use in predicting the conformational preference of chamaleon sequences
(sequences with high sequence identity and different folds). This sequence dependent conformational
preferences and per residue free energy decomposition set the stage for the use of this method in
protein design.

@@

\begin{table}
\begin{center}
\caption{The confinement method assigns a more favorable free energy to the experimentally determined structure than to
    computer-generated predictions. For each target, we examined as many as five predictions submitted by CASP
    participants. We report the free energy difference between the most favorable decoy and the experimentally
    determined structure. Positive $\Delta\Delta G (=G_{best~decoy} - G_{native})$ values indicate that the experimental structure is predicted to be
    more favorable than any of the decoys.}
\label{table:casp_control}
\begin{tabular}{l l l l}\hline
    CASP Target  & PDB Identifier & $\Delta \Delta G$ (kcal/mol) & GDT-TS \\ \hline
     T0531       &    2KJX        &          $11.15 \pm 0.70$    &  $44$ \\ \hline
     T0538       &    2L09        &          $-3.00 \pm 0.47$    &  $96$ \\ \hline
     T0540       &    3MX7        &          $16.94 \pm 0.49$    &  $70$ \\ \hline
     T0559       &    2L01        &          $2.10 \pm 0.24$     &  $94$ \\ \hline
     T0560       &    2L02        &          $18.00 \pm 0.49$    &  $94$ \\ \hline
     T0569       &    2KYW        &          $20.01 \pm 0.69$    &  $78$ \\ \hline
\end{tabular}
\end{center}
\end{table}
\Alberto{Should we comment on T0560? Why is it giving such a high value with such high gdt? It's not
the point to make here, but it is what jumps at me.}

@@

We also tested the confinement method predictions for difference free energies on the chameleon sequences of Orban et
al.\cite{Alexander2007,He2008,Alexander2009,Bryan2010,He2012}. These are instances in which two highly
similar sequences fold into remarkably different structures. Orban and co-workers have designed a protein-G-like
sequence of 56-residues that is marginally stable in one of two possible folds. By mutating key residues in this
sequence they are able to stabilize one fold or the other (see Figure~\ref{fig:orban}). We refer to the $4\beta +
\alpha$ structure as the $\beta$ conformation, and we refer to the $3\alpha$ structure as the
$\alpha$ conformation. We denote sequences that prefer the $\alpha$ fold as GA and sequences that
prefer $\beta$ as GB. One
pair of sequences (GA88/GB88) is 88 percent identical in sequence, differing in seven positions. Another pair
(GA95/GB95) is 95 percent identical, differing in three positions. Accurately predicting the structural preferences of
these structures presents a serious challenge for computational
methods \cite{Allison2011}.

@@

As is common in the CASP experiment, we assess our results in terms of Global Distance Test Total Score (GDT-TS)
\cite{Zemla2003}, which is a C$\alpha$ based measure of structural accuracy. It can be understood roughly as the
percentage of residues that are correctly positioned in the model (range 0 to 100, higher is
better). We did not have enough
computer power to analyze every model, so the initial models for the test were chosen from a selection of different
server groups that have done well in past CASP events.

@@

First, we examined the ability of the confinement method to rank different models for a target that
has been generated
by a single group using the same methodology for all predictions. We examined two targets: T0559 and T0560 (see Table~\ref{table:casp_testcase}).

\begin{table}
\begin{center}
\caption{The six different CASP targets that are used in this study. The corresponding PDB Identifier and the 
         organism type are also listed.}
\label{table:casp_testcase}
\begin{tabular}{l l l}\hline
    CASP Target  & PDB Identifier &  Organism type  \\ \hline
     T0531       &    2KJX        &  extracellular domain of the jumping  \\
                 &                &  translocation breakpoint protein  \\ \hline
     T0538       &    2L09        &  protein asr4154 from Nostoc sp. PCC7120     \\ \hline
     T0540       &    3MX7        &  human Fas apoptotic inhibitory molecule     \\ \hline
     T0559       &    2L01        &  protein BVU3908 from Bacteroides vulgatus   \\ \hline
     T0560       &    2L02        &  protein BT2368 from Bacteroides  \\ 
                 &                &  thetaiotaomicron         \\ \hline
     T0569       &    2KYW        &  domain of adhesion exoprotein from \\ 
                 &                &   Pediococcus pentosaceus \\ \hline
\end{tabular}
\end{center}
\end{table}

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=3.5 in]{T0540.pdf}
    \end{center}
    \caption{The confinement method can correctly rank the native structure and two models submitted by different
        prediction groups.}
\label{fig:T0540}
\end{figure}
\Alberto{if the drop down bars/point are a different color it will be easier to see when the crystal
is on top of the axis}
\subsection*{The per-residue free energy is sensitive to small changes in protein conformation:}

@@

The confinement method clearly identifies two hydrophobic residues V59 and I61, which destabilize the predicted model
with respect to the experimental structure (Figure~\ref{fig:T0569_per_residue} and Figure S3). The side chains of
these hydrophobic residues are oriented towards the protein hydrophobic core in the native NMR structure but oriented
towards the exterior of the protein and are solvent exposed in the model. These residues are part of a beta-sheet
in the experimental structure, but because of their side chain orientation the corresponding
beta-sheet becomes disordered in the predicted model 
(Figure~\ref{fig:T0569_per_residue} and Figure S4). There is also a large
difference around K76, which forms a salt-bridge with D11 in the predicted model, but not in the experimental structure.
This suggests that salt-bridge interactions are too favorable for the combination of force field and implicit solvent
model we use, which has been a problem noted in the past \cite{Roe2007}.

@@

Despite success for most of the studied systems, there are few failures, specially when the GDT-TS scores of the
compared structures are very close. One example is Target T0538, where we compared the experimental structure with three
models (Model 1: ``PconsR''---GDT-TS=96; Model 2: ``Shell''---GDT-TS=90; Model 3: ``FOLDIT''---GDT-TS=86). Contrary to
our expectation, the confinement method predicts that computer generated Model 1 is more stable than the crystal structure (Figure S5).
Per-residue free energy calculations (not shown) show that despite only small variations at the backbone level, the
side chains are oriented in very different ways (Figure~\ref{fig:T0538compare}), giving rise to large differences in the stabilization of
certain residues. In particular, some of the differences arise from different salt bridge patterns and certain flexible
polar residues exposed to the surface. This unexpected result shows that the confinement method is very sensitive to
local interactions (including side chain reorientation) and may indicate issues with the forcefield and implicit solvent
models used in the calculation.

\begin{figure}
\begin{center}
\includegraphics[width=4.9 in,height=4.0 in]{Target_538_compare.pdf}
\end{center}
\caption{The per residue free energy calculation found some differences of the side chain orientation in computer generated model 1 and 
experimental structure of CASP target T0538. 
(A) Two salt bridges between residues E35-R32 and E28-K24 exist in the native structure. These are
absent in the computer generated model structure. Instead
(B) a new salt bridge is formed between residues R32-E28 in model. (C) Another salt bridge between
residues  R26-E50 in the experimental structure. (D) In comparison to the native 
structure, orientation of F51 differs in the model.}
\label{fig:T0538compare}
\end{figure}

@@

A part of the CASP experiment is dedicated to the quality assessment (QA) of predicted models \cite{Kryshtafovych2011}.
Here, predictors were asked to score each model on a scale (known as qmode) from 0 to 1, with higher values
corresponding to better models \cite{Kryshtafovych2011}. Most of the groups use consensus strategies in such experiments.
Here, we chose two computer generated models from CASP taget T0538, where the top performing group "MUFOLD-WQA" \cite{Wang2011} failed.
One of the model was from "PconsR" (GDT-TS = 83, qmode 1 = 0.5865) and another from "MULTICOM-NOVEL" ( GDT-TS = 83, qmode 1 = 0.5865). We 
choose these two models as the model by PconsR was the most accurate
predicted model for this target, whereas the other model was predicted best by QA test of MUFOLD-WQA. The calculated
free energy using confinement method indicate that the model by PconsR is more stable by $3.9
Kcal/Mol$ which supports
the GDT\_TS trend.  Although consensus prediction strategies do well specially when dealing with large numbers of structures, they 
can miss a good answer that is only predicted by a few methods. At least in this case, the confinement
method is able to capture this structure that was otherwise missed.

@@

The confinement method has been described in details by Tyka et al. in reference \cite{Tyka2006} and
Cecchini et  al. \cite{Cecchini2009}. The basic approach of the confinement approach is the same in
both these papers. A thermodynamic cycle is used to compute the free energy between conformations A
and B. However, there are some small technical differences between the two approaches. Here we briefly
describe the procedure that we used.

@@

\item  In the first step, minimizations of A and B conformations are performed. These minimized conformations
    ($A^{\ast}$ and $B^{\ast}$) are the reference conformation of that state.

\item The free energy of confining the ensemble (A or B) to a tighter ensemble ($A^\ast$ or $B^\ast$) is calculated.
    This is done by gradually confining each atom to its position in the reference conformation ($A^{\ast}$ or
    $B^{\ast}$) using a series of progressively stronger position restraints. This is done by running 21 molecular
    dynamics simulation (each 20 ns long) for each leg of the thermodynamic cycle, where the harmonic restraint force
    constant was scaled from 0.00005 Kcal/Mol (mostly free) to 81.92 kcal/mol (tightly restrained). The free energy for
    this step is estimated using a thermodynamic integration approach developed by Tyka et. al. \cite{Tyka2006}. The
    confinement free energies calculated in this step are denoted as $\Delta G_{A,A^\ast}$ and $\Delta G_{B,B^\ast}$ in
    Figure~\ref{fig:method}.

@@

One advantage of confinement method is that
none of the simulations during the restraining step depends on each other. Therefore, with available computer resources it can be fast 
to compute. The computational cost for the $56$-residue protein with chameleon sequence that is reported in this article is only $4$ hours 
in a single gpu computer\Alberto{all the different confinements in one GPU?}. All calculations were
performed with the Amber 11 suit of programs~\cite{Case2012,Goetz2012} in combination with the ff99SB
forcefield~\cite{Hornak2006} and the generalized born implicit solvent \cite{Mongan2006}.

Interestingly, we extend the
method for calculation of per residue free energy in an approximate way. For this purpose, the confinement energy,
$\Delta G_{A,A*}$ and $\Delta G_{B,B*}$ of each residue is calculated in the usual numerical way as described by
Tyka et al. in reference \cite{Tyka2006}. The internal energy of each residue is calculated using
amber's "decomp" module using the
final restrained trajectory. We call this method approximate as we ignore the entropic contribution from the normal mode
or quasiharmonic analysis. However, this contribution to the total free energy is much smaller, which allow us to study
the mechanistic details of conformational preference of each residue.

### 2013-05-05 Paper/supporting_information.tex
Revisions on manuscript and SI may 5, Alberto

\section{Preparation of input model for chamelon sequences}

We prepared computer generated models in both $\alpha$ and $\beta$ conformations for all five chamelon sequences mentioned 
in the main text. Crystallographic structure of GA95 and GB95 sequence have $\alpha$ and $\beta$ fold respectively. When
we attempt to prepare the $\alpha$ and $\beta$ fold with GA95 sequence, we start with the crystallographic 
structure of GA95 and GB95 with GA95 sequence. However, we remove the sidechains at three mutated positions in both the fold 
and call the program SCWRL4 \cite{Krivov2009} to generate the sidechain conformation of the mutated residues and its neighbors. Followed by 
this step, we do long molecular dynamics
minimization to remove any possible bad contact. This procedure was followed to generate all $\alpha$ and 
$\beta$ conformations of five sequences that are used for free energy calculation.

\setcounter{figure}{0}
\makeatletter 
\renewcommand{\thefigure}{S\@arabic\c@figure}
\begin{figure}
\includegraphics[width=5.6 in,height=5.3 in]{orban_perres_SI.pdf}
\label{fig:orban_full}
\caption{Per residue free energy calculation indicate significant differences of sidechain orientation of key amino acid residues. These amino acid
residues control the free energy equilibrium to either $\beta$ or
$\alpha$ conformation. All figures in the left are in $\beta$ conformation, whereas the right ones are in $\alpha$ conformation except in Figure (F) 
where both are in $\alpha$ conformation.
Sidechains of hydrophobic residues L7 in (A) and A26 in (B) are oriented towards the protein interior in $\beta$ 
conformation. In such conformation they have hydrophobic interactions with other hydrophobic residues colored in white. In $\alpha$ conformation these residues are 
exposed to the protein surface.  
(C) Y45 forms H-bond with D47 with GB95 sequence and $\beta$ conformation. (D) Q11 and E15 forms a salt bridge in $\beta$ conformation which is absent
in the $\alpha$ conformation. (E) Hydrophobic I49 is exposed to the protein surface in $\beta$ conformation, whereas in $\alpha$ conformation it 
is inside protein hydrophobic core. (F) Sidechain of I30 is inside the protein hydrophobic core as it has a smaller sidechain with GA95 sequence 
and $\alpha$ conformation. In contrast, Phe-30 with larger sidechain is exposed to the protein surface with GB95 sequence and $\beta$ conformation.}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=5 in,height=4 in]{T0569_perres2.pdf}
\end{center}
\caption{Key differences between experimental structure and the computer generated model model as indicated by per residue free energy calculation.
The sidechain of hydrophobic residues, V59 and I61 are oriented towards the protein hydrophobic core in (A)  
structure, (B) exposed to the surfure in model. The beta sheet containing these residues is disordered in the generated model.
A salt bridge between K76 and D11 in (D) computer generated structure, which is absent in the (C) native structure.}
\label{fig:T0569_per_residue}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{T0538.pdf}
\end{center}
\caption{The native and three model structure of engineered protein from Asr4154 protein (PDB ID: 2L09 and CASP
    code:T0538). The model 1,2 and 3 are from the group PconsR, Shell and FOLDIT respectively.}
\label{fig:T0538}
\end{figure}

\section{What can we say about low resolution models from CASP experiments?}

In the main text, we discussed how well confinement method predicted preferences when the structures are not very far from the
native. But the question remains how far from native can we go and still see that the method produces correct result. In
this section we explored this question with models from  extracellular domain of the jumping translocation breakpoint
protein (pdb id: 2KJX). Most of the group could only generate low resolution models for this CASP9 target (id: T0531) .
In our comparison, we choose five models by the group MUFOLD-MD, which was the best performing group for this target
with their best model had a GDT\_TS value of 44. The result presented in Figure~\ref{fig:T0531} shows: 1.) native is
correctly identified as expected and 2.) Surprisingly there is a high level of correlation between the GDT ranking and
the free energy ranking for model 1 and model 3, the rest three structures with GDT\_TS score less than 35  are ordered
incorrectly. It is worth to note that models 2 and 3 have the same GDT and very different free energies, meaning that
the actual ordering could change a lot \cite{Perez2012}. It is encouraging that at least the method can pick out the
best model even though it is got a low GDT score: 44.

>>>>

\section*{Preparation of input models for chamelon sequences}

We prepared computer generated models in both $\alpha$ and $\beta$ conformations for all five chamelon sequences mentioned 
in the main text. The crystallographic structure of GA95 and GB95 sequences have $\alpha$ and $\beta$ fold respectively.
To prepare the $\alpha$ and $\beta$ fold for the GA95 sequence, we start with the crystallographic
conformation of the native GA95 and GB95 sequences. 
We then remove the sidechains at three mutated positions in both folds
and call the program SCWRL4 \cite{Krivov2009} to generate the side chain conformations of the
mutated residues and their neighbors. Followed by 
this step, we perform extensive molecular dynamics
minimizations to remove any possible bad contact. This procedure was used to generate all $\alpha$ and 
$\beta$ conformations for the five sequences that were used in the free energy calculation.

\setcounter{figure}{0}
\makeatletter 
\renewcommand{\thefigure}{S\@arabic\c@figure}
\begin{figure}
\includegraphics[width=5.6 in,height=5.3 in]{orban_perres_SI.pdf}
\label{fig:orban_full}
\caption{Per residue free energy calculations indicate significant differences of side chain orientation of key amino acid residues. These amino acid
residues displace the free energy equilibrium to either $\beta$ or
$\alpha$ conformation. All figures in the left are in $\beta$ conformation, whereas the right ones are in $\alpha$ conformation except in Figure (F) 
where both are in $\alpha$ conformation.
Sidechains of hydrophobic residues L7 in (A) and A26 in (B) are oriented towards the protein interior in $\beta$ 
conformation. In such conformation they have hydrophobic interactions with other hydrophobic residues colored in white. In $\alpha$ conformation these residues are 
exposed to the protein surface.  
(C) Y45 forms H-bond with D47 with GB95 sequence and $\beta$ conformation. (D) Q11 and E15 forms a salt bridge in $\beta$ conformation which is absent
in the $\alpha$ conformation. (E) Hydrophobic I49 is exposed to the protein surface in $\beta$ conformation, whereas in $\alpha$ conformation it 
is inside protein hydrophobic core. (F) Sidechain of I30 is inside the protein hydrophobic core as it has a smaller sidechain with GA95 sequence 
and $\alpha$ conformation. In contrast, Phe-30 with larger sidechain is exposed to the protein surface with GB95 sequence and $\beta$ conformation.}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=5 in,height=4 in]{T0569_perres2.pdf}
\end{center}
\caption{Key differences between the experimental structure and the computer generated models as
    indicated by the per residue free energy calculation.
The side chains of hydrophobic residues, V59 and I61 are oriented towards the protein hydrophobic core in (A)  
, in (B) they are exposed to the surface. The beta sheet containing these residues is disordered in
the generated model due to
a salt bridge between K76 and D11 in (D) the computer generated structure, which is absent in the (C) native structure.}
\label{fig:T0569_per_residue}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{T0538.pdf}
\end{center}
\caption{The native and three model structures for an engineered protein (PDB ID: 2L09 and CASP
    code:T0538). Models 1,2 and 3 are from the group PconsR, Shell and FOLDIT respectively.}
\label{fig:T0538}
\end{figure}

\section*{What can we say about low resolution models from CASP experiments?}

In the main text, we discussed how well confinement method predicted preferences when the structures are not very far from the
native. But the question remains how far from native can we go and still see that the method produces correct result. In
this section we explored this question with models from the extracellular domain of the jumping translocation breakpoint
protein (pdb id: 2KJX). Most of the groups could only generate low resolution models for this CASP9 target (id: T0531).
In our comparison, we chose five models by the MUFOLD-MD group, which was the best performing group for this target
even though their best model had a GDT\_TS value of only 44. The results presented in Figure~\ref{fig:T0531} show: 1.) native is
correctly identified as expected and 2.) Surprisingly there is a high level of correlation between the GDT ranking and
the free energy ranking for model 1 and model 3, the other three structures with GDT\_TS score less than 35  are ordered
incorrectly. It is worth to note that models 2 and 3 have the same GDT and very different free energies, meaning that
the actual ordering could change a lot \cite{Perez2012}. It is encouraging that at least the method can pick out the
best model even though it is got a low GDT score: 44.
